---
title: "EEG forward models: from a source space to a validated leadfield"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EEG forward models: from a source space to a validated leadfield}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

`PhysioHeadModels` builds the EEG *forward* operator - the leadfield that maps
cortical dipole sources to scalp potentials. This vignette walks the whole task
end to end using only the analytic and boundary-element solvers, so everything
below runs offline with no downloads.

Every leadfield here follows the ecosystem convention: an
`n_electrodes x (n_sources * 3)` matrix, with the three orientation (x, y, z)
columns stored contiguously per source.

```{r setup}
library(PhysioHeadModels)
```

## 1. A head surface and an electrode montage

A subdivided icosahedron gives a nearly uniform triangulated sphere. We use it
both as the conducting surface for the boundary-element solver and as a place to
put electrodes (here, a handful of its vertices).

```{r mesh}
mesh <- icosphere(subdivisions = 2)      # 320 triangles, 162 vertices
c(vertices = nrow(mesh$vertices), faces = nrow(mesh$faces))

electrodes <- mesh$vertices[1:19, ]      # 19 scalp electrodes on the unit sphere
```

## 2. A cortical source space

`sphericalSourceSpace()` lays sources out on an inner shell and returns the
outward radial normal at each one - the orientation a surface-constrained
forward model assumes.

```{r sources}
ss <- sphericalSourceSpace(n_sources = 60, radius = 0.6)
dim(ss$positions)
dim(ss$normals)
```

## 3. The analytic single-sphere leadfield

For a homogeneous sphere there is a closed-form (Legendre-series) solution. It
is fast and exact for this geometry, so it doubles as a reference when checking
numerical solvers.

```{r analytic}
lf <- sphereLeadfield(electrodes, ss$positions)
dim(lf)                                   # 19 electrodes x (60 sources * 3)
```

If each source's orientation is fixed to its surface normal, collapse the three
orientation columns into one with `constrainOrientation()`:

```{r constrain}
lf_fixed <- constrainOrientation(lf, ss$normals)
dim(lf_fixed)                             # 19 electrodes x 60 sources
```

## 4. A boundary-element leadfield, and validating it

`bemLeadfield()` solves the same problem numerically with a double-layer
collocation BEM, which is what you would use for a *realistic* (non-spherical)
head. On a sphere mesh we can score it against the analytic answer with
`leadfieldRDM()`: the relative difference measure (RDM) captures topography
error (0 = identical), while the magnitude ratio (MAG) captures amplitude.

```{r bem}
few_sources <- rbind(c(0, 0, 0.5), c(0.3, 0, 0.2), c(0, 0.35, 0.1))

bem <- bemLeadfield(mesh, conductivities = c(0.33, 0), electrodes, few_sources)
ana <- sphereLeadfield(electrodes, few_sources)

leadfieldRDM(bem, ana)
```

The topography (RDM) is already close and *improves as the mesh is refined*,
while the amplitude (MAG) carries the well-known coarse-collocation bias:

```{r convergence}
rdm_at <- function(sub) {
  m <- icosphere(sub)
  e <- m$vertices
  leadfieldRDM(bemLeadfield(m, c(0.33, 0), e, few_sources),
               sphereLeadfield(e, few_sources))
}
rbind(`icosphere(1)` = rdm_at(1),
      `icosphere(2)` = rdm_at(2))
```

Because source localization is invariant to a global leadfield scale, RDM is the
metric to trust for validation; the absolute amplitude bias does not affect the
inverse solution.

## Where to go next

- Published, precomputed head models (the New York Head, FreeSurfer fsaverage)
  are fetched on demand with `fetchNYHead()` / `nyHeadLeadfield()` and
  `fetchFsaverage()`; they are downloaded to `headModelCache()` rather than
  shipped, and the New York Head reader needs the **hdf5r** package.
- The leadfields and source spaces produced here feed EEG signal processing and
  inverse-analysis workflows in the **PhysioEEG** package.

Forward-model geometry, coordinate frames, conductivity assumptions, and the
reference convention should always be checked against the needs of each study.
